Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-03-21 text/paper2/ms.tex
acknowledgements

\newcommand{\D}{{\bf D}}
\newcommand{\W}{{\bf W}}
\newcommand{\X}{{\bf X}}
\newcommand{\J}{{\boldsymbol J}}
\newcommand{\bSigma}{{\bf \Sigma}}
\newcommand{\bsigma}{\boldsymbol\sigma}
\newcommand{\rtide}{r_{\rm tide}}
\newcommand{\bs}{\boldsymbol}
\newcommand{\sat}{{\rm p}}
\newcommand{\tub}{t_{\rm ub}}
\newcommand{\tint}{t_{\rm int}}
\newcommand{\rr}{\widetilde{\bs{r}}}
\newcommand{\vv}{\widetilde{\bs{v}}}
\newcommand{\tailbit}{\beta}
\newcommand{\Loffset}{\alpha}
\newcommand{\pshock}{P_{\rm shock}}
\newcommand{\paperone}{Paper 1}
\newcommand{\vhalo}{v_{\rm h}}

@@

%[][hereafter \paperone]
In \citet{apw13}, we introduced a simple method (\emph{Rewinder}) for using individual stars associated with tidal debris combined with knowledge about the mass and orbit of the progenitor to constrain properties of the host galaxy potential. The method exploits the relationship between the phase-space distribution of debris and (measurable) properties of the progenitor system (e.g., the tidal radius and escape velocity).  Specifically, a better model for the potential is one in which orbits of test particle stars (integrated backwards from their present position) came close (within the tidal radius in position, and escape velocity in velocity) to the orbit of the progenitor. In this article, we use these simple scalings --- which depend only on the mass and orbit of the progenitor and parent potential --- to present a fully probabilistic method for modeling tidal streams in any underlying potential given data with realistic uncertainties. The method relies on numerical orbit integration in ordinary phase-space and can thus incorporate arbitrarily complex models for the parent potential. The model uses individual stars as constraints --- not density matching --- and can therefore handle very small samples of well-measured stars (e.g., 4-16). The model trivially extends to incorporating multiple streams, debris structures, and other kinematic information.

@@

We leave the potential parameters ($q_1$, $q_z$, $\phi$, $\vhalo$) free to vary, along with the Lagrange point offset, $\Loffset$, and initialize an ensemble of 64 MCMC walkers by sampling from the priors summarized in Table~\ref{tbl:params}. We run the walkers for 10000 steps to burn-in, then restart the sampler starting from the final position of the burn-in phase and run for another 5000 steps. Figure~\ref{fig:trace} shows the walker positions over the 5000 inference steps for each of the five parameters. The autocorrelation time for each parameter is displayed on its corresponding panel and the horizontal lines indicate the true simulation values for the potential parameters. Figure~\ref{fig:exp1_posterior} shows projections of the posterior probability distribution for the parameters; with perfect data, the uncertainties on the potential parameters are all $<1\%$.

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\textwidth]{../../plots/paper2/mcmc_trace.png}
\caption{ TODO: mcmc trace plots  }\label{fig:trace}
\end{center}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\textwidth]{../../plots/paper2/exp1_posterior.png}
\caption{ TODO  }\label{fig:exp1_posterior}
\end{center}
\end{figure}

\subsection{Data with near-future uncertainties}
We next take the same eight stars used in the previous experiment and ``observe'' them with optimistic observational uncertainties. We take these stars to be RR Lyrae variables which are known to be excellent distance indicators via the mid-infrared period-luminosity relation \citep[as shown in, e.g.,][]{madore12}. $\sim$100 RR Lyrae associated with the Sgr stream and $\sim$30 associated with the Orphan stream will be observed with \spitzer as part of the SMASH survey \citep{smashprop} with expected fractional distance uncertainties around $\sim$2\%. These stars will also be included in the \gaia\, proper motion catalog: at a distance of $\sim$50~kpc, a typical RR Lyrae will have a tangential velocity error around $\sim$20~km/s.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\textwidth]{../../plots/paper2/exp2_particle0.png}
\caption{ TODO  }\label{fig:exp2_particle0}
\end{center}
\end{figure}

@@

\acknowledgements
Rix, Bovy\\
Gaia data challenge / Surrey / Gaia data team / Anthony Brown \\ % check with AB about citations
APW is supported by a National Science Foundation Graduate Research
Fellowship under Grant No.\ 11-44155. This work was supported in part by 
the National Science Foundation under Grant No. PHYS-1066293 and the 
hospitality of the Aspen Center for Physics.\\
DWH was partially supported by NASA (Grant NNX12AI50G) and the NSF (Grant IIS-1124794).\\
This research made use of Astropy, a community-developed core Python package for Astronomy \citep{astropy13}. \\
Columbia HPC\\

>>>>

\newcommand{\D}{{\bf D}}
\newcommand{\W}{{\bf W}}
\newcommand{\X}{{\bf X}}
\newcommand{\J}{{\boldsymbol J}}
\newcommand{\bSigma}{{\bf \Sigma}}
\newcommand{\bsigma}{\boldsymbol\sigma}
\newcommand{\rtide}{r_{\rm tide}}
\newcommand{\bs}{\boldsymbol}
\newcommand{\sat}{{\rm p}}
\newcommand{\tub}{t_{\rm ub}}
\newcommand{\tint}{t_{\rm int}}
\newcommand{\rr}{\widetilde{\bs{r}}}
\newcommand{\vv}{\widetilde{\bs{v}}}
\newcommand{\tailbit}{\beta}
\newcommand{\Loffset}{\alpha}
\newcommand{\pshock}{P_{\rm shock}}
\newcommand{\paperone}{Paper 1}
\newcommand{\vhalo}{v_{\rm h}}
\newcommand{\rewinder}{\emph{Rewinder}}
\newcommand{\potp}{$q_1,q_z,\phi,\vhalo$}

@@

%[][hereafter \paperone]
In \citet{apw13}, we introduced a simple method for using individual stars associated with tidal debris combined with knowledge about the mass and orbit of the progenitor to constrain properties of the host galaxy potential. The method exploits the relationship between the phase-space distribution of debris and (measurable) properties of the progenitor system (e.g., the tidal radius and escape velocity).  Specifically, a better model for the potential is one in which orbits of test particle stars (integrated backwards from their present position) came close (within the tidal radius in position, and escape velocity in velocity) to the orbit of the progenitor. In this article, we use these simple scalings --- which depend only on the mass and orbit of the progenitor and parent potential --- to present a fully probabilistic method (\rewinder) for modeling tidal streams in any underlying potential given data with realistic uncertainties. By ``model'' here we mean a justified, parametrized likelihood function with priors on the parameters. \rewinder\, relies on numerical orbit integration in ordinary phase-space and can thus incorporate arbitrarily complex models for the parent potential. Individual stars are used as constraints, thus \rewinder\, can handle very small samples of well-measured stars (e.g., 4-16). This model trivially extends to incorporating multiple streams, debris structures, and other kinematic information.

@@

We leave the potential parameters ($q_1,q_z,\phi,\vhalo$) free to vary, along with the Lagrange point offset, $\Loffset$, and initialize an ensemble of 64 MCMC walkers by sampling from the priors summarized in Table~\ref{tbl:params}. We run the walkers for 10000 steps to burn-in, then restart the sampler starting from the final position of the burn-in phase and run for another 5000 steps. Figure~\ref{fig:trace} shows the walker positions over the 5000 inference steps for each of the five parameters. The autocorrelation time for each parameter is displayed on its corresponding panel and the horizontal lines indicate the true simulation values for the potential parameters. Figure~\ref{fig:exp1_posterior} shows projections of the posterior probability distribution for the parameters; with perfect data, the uncertainties on the potential parameters are all $<1\%$.

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=0.9\textwidth]{../../plots/paper2/mcmc_trace.png}
\caption{ Positions of all 64 MCMC walkers (see Section~\ref{sec:experiments}) for each parameter at each step of the inference (black lines). Horizontal green lines show the input values of the potential parameters (\potp). The MCMC chains do not display any bulk movement or stray walkers, implying, by eye, that the chains have converged. Projections of the sample densities are shown in Figure~\ref{fig:exp1_posterior}. }\label{fig:trace}
\end{center}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\textwidth]{../../plots/paper2/exp1_posterior.png}
\caption{ Projections of the posterior probability distribution over the four potential parameters (\potp) and Lagrange point offset ($\alpha$) assuming negligible uncertainties on the observed phase-space coordinates of eight stars and the progenitor, visualized as two-dimensional histograms of MCMC samples. Solid contours (black lines) show approximately 1$\sigma$ and 2$\sigma$ levels of the distribution. Vertical and horizontal lines (blue) show the true, input values for the potential parameters used in the n-body simulations. For the potential parameters, the axis ranges were chosen to be $\pm$5\% around the true values. Note that any biases in the inferred values are $\lesssim1\%$. }\label{fig:exp1_posterior}
\end{center}
\end{figure}

\subsection{Data with near-future uncertainties}
We next take the same eight stars used in the previous experiment and ``observe'' them with optimistic observational uncertainties. We take these stars to be RR Lyrae variables which are known to be excellent distance indicators via the mid-infrared period-luminosity relation \citep[as shown in, e.g.,][]{madore12}. Nearly 100 RR Lyrae associated with the Sgr stream and $\sim$30 associated with the Orphan stream will be observed with \spitzer as part of the SMASH survey \citep{smashprop} with expected fractional distance uncertainties around $\sim$2\%. These stars will also be included in the \gaia\, proper motion catalog: at a distance of $\sim$50~kpc, a typical RR Lyrae will have a tangential velocity error around $\sim$20~km/s.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\textwidth]{../../plots/paper2/exp2_particle.png}
\caption{ TODO  }\label{fig:exp2_particle}
\end{center}
\end{figure}

@@

\acknowledgements
We thank Anthony Brown (Leiden) for insight on the \gaia\, data quality and for providing the open source \texttt{PyGaia} code for computing predicted uncertainties. We thank the organizers of the Gaia Challenge (2013) and the University of Surrey for hospitality during this workshop. We thank Hans-Walter Rix (MPIA) and Jo Bovy (IAS) for ... [TODO: HOGG].
APW is supported by a National Science Foundation Graduate Research Fellowship under Grant No.\ 11-44155. This work was supported in part by the National Science Foundation under Grant No. PHYS-1066293. 
DWH was partially supported by NASA (Grant NNX12AI50G) and the NSF (Grant IIS-1124794).
This research made use of Astropy, a community-developed core Python package for Astronomy \citep{astropy13}.
This work additionally relied on Columbia University's \emph{hotfoot} and \emph{yeti} compute clusters, and we acknowledge the Columbia HPC support staff for assistance. \\
